\section*{Acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

Foremost, I am deeply grateful to my thesis advisor, Prof.\ Bj\"orn Sandstede,
whose influence on this project has been profound and irreplaceable.
He gave me the initial ideas that set the direction of this work, and he
remained a constant source of guidance as those ideas evolved into a full
pipeline.
Every time I brought him a result I did not understand, he found the right
question to ask; every time I was moving in the wrong direction, he saw it
before I did and redirected me with patience.
His validation --- knowing that the mathematical choices I was making were
sound --- gave me the confidence to push further than I would have dared
on my own.
His feedback on drafts was precise and direct, always pointing to what
mattered most, and his high standards for clarity and rigor are visible
throughout this text.
Beyond any specific contribution, it was his way of thinking about
mathematics and science that I have tried to absorb and carry forward.
I am also grateful to my second reader, Prof.\ Ian~Wong, for his thoughtful
reading and encouragement.

This thesis was an opportunity to see the full power of applied mathematics
and modern computation working together.
Looking back, it draws on nearly every area I studied at Brown in the
APMA--CS--Mathematics track, often in ways I did not anticipate when I
first encountered those subjects.
Probability theory and statistics underlie the KDE fields that convert
discrete agent trajectories into continuous density snapshots, the noise
models, and the convergence analysis.
Linear algebra and functional analysis provide the backbone: POD and
SVD truncation, the Hilbert-space formulation of the weak form, and the
Kolmogorov $n$-width theory that explains \emph{why} transport-dominated
systems resist linear reduction.
Differential equations are everywhere --- in the ETDRK4 integrator, in
the PDE candidate library, in the Fokker--Planck framing of the mean-field
limit.
Machine learning and deep learning enter through the LSTM sequence models
and the SR3 sparse regression framework.
Abstract algebra makes a perhaps unexpected appearance: the continuous
SO(2) rotational symmetry of the system is precisely the kind of group
action studied in algebra, and the shift-alignment step is, in essence,
quotienting the data by that group action to obtain a canonical
representative.
Topology threads through the persistent homology literature against which
we position our approach.
And statistics returns at the end in the form of stability selection and
model-order criteria.

I am grateful to all the faculty and courses at Brown that built these
foundations in me, and I find it genuinely exciting that a single
undergraduate thesis could call on all of them at once.
Applied mathematics is at its best when it refuses to stay in one lane ---
and high-performance computing gave us the experimental scale to take
those mathematical ideas seriously.

I thank the Center for Computation and Visualization
(CCV) at Brown University for providing access to
the OSCAR high-performance computing
cluster, on which all particle simulations,
ROM training, WSINDy fitting, and ETDRK4 rollout computations were
performed.
